---
title: "Full NR and LASSO Run"
author: "Tucker Morgan - tlm2152"
date: "3/26/2022"
output: pdf_document
---

This file prepares our data, runs Newton-Raphson, runs LASSO, and plots some results at the bottom. I think this file should stay clean and relatively simple. I think the file should only take 15ish minutes to knit or run in full. I think the best idea is for plots to be saved as R objects and imported into the presentation / report, but open to ideas here.

```{r setup, warning=FALSE, message = FALSE}
source("./shared_code/setup.R")
knitr::opts_chunk$set(cache = TRUE)
```

```{r data import and partition, warning=FALSE, message = FALSE}
source("./shared_code/data_prep.R")
```

# Newton-Raphson

```{r newton-raphson function}
source("./shared_code/full_NR2.R")
source("./shared_code/roc_func.R")
roc_nr <- roc_func(nr_beta_est, bc_tst)
```

# LASSO

```{r external-functions, warning=FALSE, message=FALSE}
source("./shared_code/logistic_lasso.R")
source("./shared_code/auc_calc_lasso.R")
```

## Folding Data

```{r folding training data}
source("./shared_code/cv_folding.R")

bc_trn_folds <- 
  cv_sets(training = bc_trn) %>% 
  select(-fold_p)
```

```{r implementing cv}
source("./shared_code/cv_implementation.R")
lambda_max <- max(t(scale(as.matrix(bc_trn[,-1]))) %*% bc_trn[,1]) / nrow(bc_trn[,-1])
lambda_list <- list(log(lambda_max), log(0.0001), -(log(lambda_max) - log(0.0001))/100, exp)

cv_res <- cv_jt(k = 5, training = bc_trn_folds, 
                func = logistic_lasso, lam_start_stop_func = log, 
                lambda_list = lambda_list)
```

```{r lasso results}
# pulling out key results
selected_lambda <- 
  cv_res[[1]][[length(cv_res[[1]])]] %>% 
  filter(mean_auc == max(mean_auc)) %>% 
  pull(lambda) %>% 
  mean()

lasso_final_model <- logistic_lasso(inputs = bc_trn[,-1], 
                                    output = bc_trn[,1], 
                                    lambda_vec = selected_lambda)
lasso_betas <- lasso_final_model[[2]] %>% t()
# this is just for viz purposes
tst_lambda_vec <- exp(seq(from = log(lambda_max), 
                          to = log(0.0001), 
                          by = -(log(lambda_max) - log(0.0001))/100))

lasso_final_range <- logistic_lasso(inputs = bc_trn[,-1], 
                                    output = bc_trn[,1], 
                                    lambda_vec = tst_lambda_vec)

lfr_df <- data.frame(do.call(cbind, lasso_final_range)) %>% 
  select(-selected) %>% 
  pivot_longer(cols = starts_with("beta"),
               names_prefix = "beta.",
               names_to = "beta_coef",
               values_to = "coef_est")

# storing lasso ROC object
roc_lasso <- roc_func(lasso_betas, bc_tst)

cv_res_lam <- cv_res[[1]]


```

```{r plotting some results}
# each of these plots could be saved as R objects and imported into other documents
lfr_df %>% 
  group_by(lambda) %>% 
  filter(beta_coef != "intercept") %>% 
  ggplot(x = lambda, y = coef_est, group = beta_coef) +
  geom_path(aes(x = lambda, y = coef_est, group = beta_coef, col = beta_coef)) +
  coord_cartesian(xlim = c(0, 0.05)) + # this zooms in on the plot, comment out if flipping
#  scale_x_reverse() +                  # this flips the x-axis if wanted
#  coord_cartesian(xlim = c(0.06, 0)) + # this zooms in if flipped
  geom_vline(xintercept = selected_lambda) + 
  labs(title = "Beta Coefficients in LASSO Model") 

# this looks alright, might prefer a larger lambda range to show AUC down to 0.5
# but probably not important
cv_res[[1]][[1]] %>% 
  data.frame() %>% 
  ggplot(x = lambda, y = mean_auc) +
  geom_line(aes(x = lambda, y = mean_auc), col = "dodgerblue") +
  geom_vline(xintercept = selected_lambda, linetype = "dashed", col = "red") +
  labs(title = "Mean AUC vs. Lambda",
       x = "Lambda",
       y = "Mean AUC")


# Final ROC plots
auc_vec <- c(round(roc_lasso$auc[1], digits = 4), round(roc_nr$auc[1], digits = 4))
model_names <- c("LASSO", "Newton-Raphson")
ggroc(list(roc_lasso, roc_nr), legacy.axes = TRUE) +
  scale_color_discrete(labels = paste0(model_names, " (", auc_vec, ")"),
                       name = "Models (AUC)") +
  geom_abline(intercept = 0, slope = 1, color = "grey") +
  labs(title = "ROC Curves from Test Data Predictions")

## another option for plotting ROC curves

plot(roc_lasso, legacy.axes = TRUE, col = "goldenrod", 
     main = "ROC Curves from Test Data Predictions")
plot(roc_nr, col = "dodgerblue", add = TRUE)
legend("bottomright", legend = paste0(model_names, ": ", auc_vec), 
       col = c("goldenrod", "dodgerblue"), lwd = 2)
```

```{r tabling some results}
beta_names <- 
  (dimnames(lasso_betas)[1]) %>% 
  data.frame() %>% 
  rename(beta_coef = 1)

cbind(nr_beta_est, lasso_betas) %>% 
  data.frame() %>% 
  round(digits = 4) %>% 
  rename(NewtonRaphson = 1, LASSO = 2) %>% 
  knitr::kable(caption = "Final Beta Coefficient Estimates")
```

## Comparison to GLMNET Lasso
