---
title: "Untitled"
author: "Amy Pitts"
date: "3/17/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("shared_code/setup.R")
```


## Data import 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
bc <- read.csv("data/breast-cancer.csv") %>% as_tibble()
```

This dataset has `r nrow(bc)` rows and `r ncol(bc)` columns. The outcome variable of interst is `Diagnosis` which takes on values benign or malignant cases.  There are `r table(bc$diagnosis)[[1]]` benign cases and `r table(bc$diagnosis)[[2]]` malignant cases as seen below in Table 1. One variable is `id` and the rest 30 variables are the mean, sd and largest values of the following criteria. 

- radius (mean of distances from center to points on the perimeter)
- texture (standard deviation of gray-scale values)
- perimeter
- area
- smoothness (local variation in radius lengths)
- compactness (perimeterˆ2 / area - 1.0)
- concavity (severity of concave portions of the contour)
- concave points (number of concave portions of the contour)
- symmetry
- fractal dimension ("coastline approximation" - 1)

Using this dataset models will be compared and compared on their ability to predict cancer diagnosis. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
bc %>% 
  select(-id) %>% 
  mutate(
    diagnosis = ifelse(diagnosis == "B", "Benign", "Malignant")
  ) %>%
  tbl_summary( by = diagnosis,
               statistic = list(all_continuous() ~ "{mean} ({sd})",
                                all_categorical() ~ "{n} / {N} ({p}%)"),
               digits = all_continuous() ~ 3)  %>% 
  add_p()
```

## Build a logistic model to classify the images into malignant/benign 
Before building the model we need to first write down the  likelihood function, its gradient and Hessian matrix.


The likelihood function for our data which has a single binary response and 30 numerical explanatory variables is
\[
\pi_i = P(Y_i=1|x_{i,1}, \dots x_{i,30}) = \frac{e^{\beta_0+\beta_1x_{i,1}+ \dots \beta_{30}x_{i,30}}}{1 + e^{\beta_0+\beta_1x_{i,1}+ \dots \beta_{30}x_{i,30}}} = \frac{e^{\beta_0+\sum^{30}_{j=1}\beta_ix_{i,j}}}{1 + e^{\beta_0+\sum^{30}_{j=1}\beta_ix_{i,j}}}
\]

Where $\mathbf{X_i}$ represents the $i$ observation of all 30 of our predictor variables. 
For the data give we have the likelihood is given by 
\[ L(\mathbf{X}| \mathbf{\beta}) = \prod^n_{i=1} \left[ \pi_i^{y_i}(1-\pi_i)^{1-y_i} \right]
\]
Finding the log-likelihood we have 
\[ l(\mathbf{X}| \vec{\beta}) =\sum^n_{i=1} \left[ y_i\left(\beta_0 + \sum^{30}_{j=1}\beta_ix_{i,j}  \right) - \log\left( 1 + \exp\left(\beta_0 + \sum^{30}_{j=1} \beta_ix_{i,j} \right) \right) \right]
\]

The gradient can then can be solved for. Observe 

\[ \nabla l(\mathbf{X}|\vec{\beta}) = \left[ \sum^n_{i=1} \begin{matrix} y_i-\pi_i & \sum^n_{i=1} x_{i,1}(y_i-\pi) & \dots & \sum^n_{i=1} x_{i,30} (y_i-\pi) \end{matrix}\right]^{T}_{(1 \times 31)} 
\]

Finally, with the gradient we can derive our hessian. Note that due to the 30 predictor variables the hessian will be a 31 by 31 matrix. 

\begin{align*}
 \nabla^2 l(\mathbf{X}|\vec{\beta}) &= - \sum^n_{i=1} \begin{pmatrix} 1 \\ X \end{pmatrix} \begin{pmatrix} 1 & X \end{pmatrix} \pi_i (1-\pi_i) \\
 &= - \begin{pmatrix} 1 & X \end{pmatrix} diag( \pi_i (1-\pi_i)) \begin{pmatrix} 1 \\ X \end{pmatrix}
\end{align*}

Where $X = (x_{i,1}, \dots, x_{i, 30})$. Note that this matrix will always be negative definite at all parameters making the this a well behaved problem. 



## Develop a Newton-Raphson algorithm to estimate your model

Modifications: 
- I include half stepping in the Newton-Raphson method. 

Want to include: 
- Assent direction 

```{r}
dat <- bc %>% select(-id) %>%
  rename(y = diagnosis) %>%
  mutate(y = ifelse(y=="B", 0, 1))
```


```{r}
loglike_func <- function(dat, betavec){
  dat_x = unname(as.matrix(dat[,-1]))
  u = dat_x%*%betavec
  pi <- exp(u) / (1+exp(u))
  
  # loglikelihood
  loglik <- sum(dat$y*u - log(1 + exp(u)))
  
  #gradient 
  grad <- t(dat_x)%*%(dat$y - pi)
  
  # Hessian 
  save_mat = matrix(0, nrow = dim(dat)[1], ncol = dim(dat)[1]) 
  diag(save_mat)= pi*(1-pi)
  hess = -t(dat_x)%*% save_mat %*%(dat_x)
  
  return(list(
    loglik = loglik,
    grad = grad,
    hess = hess
  ))
  
}

# loglike_func(dat, betavec = c( rep(0.03, 30))) 

```


This is not working right now.... 
```{r}
NewtonRaphson <- function(dat,  start, tol=100, maxiter = 20){
  i = 0
  cur = start 
  stuff = loglike_func(dat, cur)
  res <- c(i=0, "loglik" = stuff$loglik,  cur, "step" = 1)
  prevloglik <- -Inf # To make sure it iterates
  while(i < maxiter && abs(stuff$loglik - prevloglik) > tol) {
    step = 1
    i <- i + 1
    prevloglik <- stuff$loglik
    
    # ascent direct check 
    eigen_vals = eigen(stuff$hess)
    if(max(eigen_vals$values) <= 0 ){ # we don't want neg def 
      hess = stuff$hess
    } else{ # if it is neg def then need to adjust 
      hess = stuff$hess - (max(eigen_vals$values) + 0.1)*diag(nrow(stuff$hess))
    }
    prev  <- cur
    cur   <- prev - rep(step, length(prev))*(solve(stuff$hess) %*% stuff$grad)
    stuff <- loglike_func(dat, cur) # log-lik, gradient, Hessian
    
    # doing half stepping 
    while(stuff$loglik < prevloglik){
      stuff <- loglike_func(dat, prev)
      step = step / 2 # this is where half steping happens 
      cur <- prev - rep(step, length(prev))*(solve(stuff$hess) %*% stuff$grad)
      stuff = loglike_func(dat, cur)
    }
    # Add current values to results matrix
    res <- rbind(res, c(i, stuff$loglik, cur, step))
  }
  
  colnames(res) <- c(names(dat[,-1]), "i", "loglik", "step")
  return(res)
}
#Running the algorithm with good starting values:
#betavec = rep(1, 31)
betavec = c(rep(0.02, 30))
ans <- NewtonRaphson(dat,betavec)

data.frame(
  i = data.frame(ans)$i,
  step = data.frame(ans)$step,
  loglik = data.frame(ans)$loglik
)
   
```

